# Randomization and allocation, blinding and placebos


## Methods of allocation

### Simple randomization
Using *simple randomization* for allocation of $2n$ patients to two treatment groups, A and B, the number of patients $N_A$ in group A has a binomial distribution: $N_A \sim \Bin(2n, 1/2)$, likewise for $N_B$. However, one is fixed by the other: $N_B = 2n - N_A$. The distribution of the larger group size $N_{\max}=\max(N_A,N_B)$ can be derived as

\[
\Pr(N_{\max}=r) = \left\{
\begin{array}{rl}
 2^{-2n} {2n \choose n} & \mbox{ for } r=n \\[.25cm]
  2^{1-2n} {2n \choose r} & \mbox{ for } r=n+1,\ldots, 2n.
\end{array}
\right.
\]


```{proof}
Given a specific $N_A$, $N_{\max}$ takes the following value:
```

\begin{table}[!h]
\centering
\begin{tabular}{p{1cm} p{1cm} p{1cm} p{1cm} c c c c c c}
  \toprule
  $N_A$ & $0$ & $1$ & $\dots$ & $n - 1$ & $n$ & $n + 1$ & $\dots$ & $2n - 1$ & $2n$ \\
  $N_B$ & $2n$ & $2n - 1$ & $\dots$ & $n + 1$ & $n$ & $n - 1$ & $\dots$ & $1$ & $0$ \\
  \midrule
  $N_{\max}$ & $2n$ & $2n - 1$ & $\dots$ & $n + 1$ & $n$ & $n + 1$ & $\dots$ & $2n - 1$ & $2n$ \\
\bottomrule
\end{tabular}
\end{table}

```{r, results='asis', echo=FALSE}
library(knitr)
library(kableExtra)

# Define the table data
data <- data.frame(
  Na = c("$N_A$", "$0$", "$1$", "$\\dots$", "$n - 1$", "$n$", "$n + 1$", "$\\dots$", "$2n - 1$", "$2n$"),
  Nb = c("$N_B$", "$2n$", "$2n - 1$", "$\\dots$", "$n + 1$", "$n$", "$n - 1$", "$\\dots$", "$1$", "$0$"),
  Nmax = c("$N_{\\max}$", "$2n$", "$2n - 1$", "$\\dots$", "$n + 1$", "$n$", "$n + 1$", "$\\dots$", "$2n - 1$", "$2n$")
)

# Transpose the data to match the original table layout
data <- t(data)

# Create a data frame from the transposed data
df <- as.data.frame(data, stringsAsFactors = FALSE)
colnames(df) <- NULL

# Print the table using kable
kable(df, format = "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, position = "center")
```

Equal sample size for both groups, $r = n$, is a special case as it appears only once:
\begin{equation*}
\Pr(N_{\max} = n) = \Pr(N_A = n)
= \binom{2n}{n} \left(\frac{1}{2}\right)^{2n} = 2^{-2n} \binom{2n}{n}.
\end{equation*}

For $r = n + 1, \dots, 2n$, the larger group can be either A or B, leading to:
\begin{eqnarray*}
\Pr(N_{\max} = r)
& = & \Pr(N_A = r) + \Pr(N_A = 2n - r) \\
& = & \binom{2n}{r} \left(\frac{1}{2}\right)^{2n} +  \binom{2n}{2n - r} \left(\frac{1}{2}\right)^{2n} \\
& \overset{\text{symm.}}{=} & 2 \cdot \binom{2n}{r} \left(\frac{1}{2}\right)^{2n} 
= 2^{1 -2n} \binom{2n}{r}.
\end{eqnarray*}


For example, for $n=15$, $\Pr(N_{\max} \geq 20) = 0.10$. So, there is a substantial chance that the two groups will end up with markedly differing sizes, as illustrated by Figure \@ref(fig:simple). Unequal group sizes lead to a *loss in power*:

\[
\mbox{ Power } = \Phi\left(\sqrt{({n_A n_B})/({n_A+n_B})} - z_{1-\alpha/2}\right)
\]

For example, for Cohen's $d=1$, total sample size $2n = n_A+n_B=30$ and $\alpha=0.05$ the loss in power is illustrated in Figure \@ref(fig:powerloss).


```{r simple, fig.cap = "Probability for unequal group sizes with simple randomization and a total sample size of $2n=30$.", echo=F}
P.NMax <- function(n, r) 2^(1-2*n)*choose(2*n, r)
n <- 15
support <- c(n:(2*n))
dist.NMax <- P.NMax(n, support[-1])
dist.NMax <- c(1-sum(dist.NMax), dist.NMax)
#round(sum(dist.NMax[support>=20]),2)

mycol <- c(rep(1, 5), rep(2, 21))
par(las=1)
plot(support, dist.NMax, type="p", ylab="Probability", xlab="Size of larger group", col=mycol, pch=20)
for(i in 1:n) 
    lines(rep(14+i, 2), c(0, dist.NMax[i]), col=mycol[i])
text(24, 0.06, "Pr = 0.10", col=2)
``` 



```{r powerloss, fig.cap = "Loss in power for unequal group sizes in a scenario with $d=1$, $n_1+n_2=30$ and $\alpha=0.05$.", echo=F}
n <- 30
n1 <- c(1:(n-1))
n2 <- n-n1
power <- c(0, pnorm(sqrt((n1*n2)/(n1+n2))-1.96), 0)
n1 <- c(0, n1, n)
par(las=1)
plot(n1, power*100, type="b", xlab="Group size in group A", ylab="Power (in %)", cex=0.65, pch=19, ylim=c(0,80))
lines(c(-2,32), rep(max(power*100), 2), col=2, lty=2)
lines(c(-2,32), rep(60, 2), col=1, lty=2)
lines(c(-2,32), rep(40, 2), col=1, lty=2)
lines(c(-2,32), rep(20, 2), col=1, lty=2)
text(26.5, 73, labels=paste("max power = ", as.character(round(max(power*100))), "%", sep=""), col=2)
``` 


### Block randomization
The problem of unbalanced group sizes can be solved by using a form of restricted randomization with so-called *random permuted blocks* (RPBs). For example, with blocks of length 4, there are the following six different sequences of length 4 that comprise two As and two Bs:

```{r, results='asis', echo=FALSE}

library(knitr)
library(kableExtra)
data <- data.frame(
  row_num = c(1, 2, 3, 4, 5, 6), 
  col1 = c("A", "A", "A", "B", "B", "B"),
  col2 = c("A", "B", "B", "B", "A", "A"),
  col3 = c("B", "B", "A", "A", "A", "B"),
  col4 = c("B", "A", "B", "A", "B", "A"))

kable(data, format = "html", escape = FALSE, col.names = NULL) %>%
  kable_styling(full_width = FALSE, position = "center")
```

In this case, the randomization selects randomly a block for every group of four patients. Block randomization using RPBs of length 4 ensures that group sizes never differ by more than 2. After every fourth patient, the two treatment groups must have the same size.

RPBs can also be used with other block lengths or for more than two groups, \eg for a block length of 6 and three groups:
```{r, echo = FALSE, message = FALSE}
library(knitr)
library(kableExtra)

# Define the table data
df <- data.frame(
  row_num = c(1, 2, 3, "$\\vdots$", 90),
  col1 = c("A", "A", "A", "$\\vdots$", "C"),
  col2 = c("A", "B", "B", "$\\vdots$", "C"),
  col3 = c("B", "B", "A", "$\\vdots$", "B"),
  col4 = c("B", "A", "B", "$\\vdots$", "B"),
  col5 = c("C", "C", "C", "$\\vdots$", "A"),
  col6 = c("C", "C", "C", "$\\vdots$", "A")
)

# Print the table using kable
kable(df, format = "html", escape = FALSE, col.names = NULL) %>%
  kable_styling(full_width = FALSE, position = "center")
```


```{example}
\code{R} code for RPBs of length 4:
```

```{r echo=TRUE, message = FALSE}
library(randomizeR)
## bc: length of block
obj4 <- pbrPar(bc=4)
getAllSeq(obj4)
```

:::{.example #ulcers}
Am RCT was conducted to examine the use of cold atmosperic plasma 
as a method to treat diabetic foot ulcers [@Mirpour2020]. Statement from the 
Methods section: 
"Patients were randomly, double-blinded, assigned to an SC groupe or SC + CAP
using bloc randomization with mixing block sizes of $4$. The size of blocks
was not disclosed for minimizing the chance of cracking the code. 
An investigator with no clinical involvement in the trial prepared a
computer random number list for block randomization. A trained physician and 
nurse who were blinded to the randomization method and treatment assignment 
collected the data. The data also were analyzed by a blinded
investigator to the study groups."
:::

In some instances, knowledge of the block length and of previous treatments
allows the next treatment to be predicted. This may cause *selection bias*.
RPBs with *random block length* have been suggested to avoid the potential for
selection bias.

```{example}
R code for RBPs with random block length 2 and 4:
```


```{r echo=TRUE}
obj24 <- rpbrPar(N=100, rb=c(2,4))
(seq24 <- genSeq(obj24))

# Randomization list
as.character(getRandList(seq24))
``` 



### Unequal randomization
Unequal allocation can provide greater experience of a new treatment and may 
even encourage recruitment in certain trials. Provided the imbalance is no
greater than 2:1, the loss in power is small (\eg from 90\% to 86\% for 
$\alpha=5\%$).
%% RPBs could easily be adapted to provide a means of ensuring an imbalance 
close to 2:1. 
Treatment allocation sequences could be built by randomly selecting from the 
15 blocks of length 6, comprising 4 As and 2 Bs.

```{example}
R code for 2:1 allocation ratio using RPBs of length 6 for two groups:
```

```{r echo=TRUE}
library(randomizeR)
## bc: length of each block
obj <- pbrPar(bc=6, ratio=c(2,1))
genSeq(obj, r=3, seed=123)
``` 


### Stratification
*Stratification* is useful when there is imbalance with respect to prognostic factors. Suppose we wish to compare a new treatment $T$ with placebo $P$ to see if it improves a certain continuous outcome. A RCT is conducted with $n$ patients in each group. %%  (perhaps using RPBs).
An important binary *prognostic factor* (\eg age) has level $A$ (child) in a proportion %% $\theta$ 
of the eligible patients, otherwise it has level $B$ (adults).
Suppose that there is *no treatment* effect, but the mean outcomes $\mu_A$ and $\mu_B$ for type $A$ and $B$ differ: $\mu_A \neq \mu_B$.
We have conducted one particular trial with $n$ patients per treatment group and 
%
\begin{itemize}  
-   $n_T$ patients of type $A$ in group $T$ and
-   $n_P$ patients of type $A$ in group $P$.
\end{itemize}

The expected 
outcome difference between group $T$ and $P$ then is

$$(n_T-n_P)(\mu_A-\mu_B)/n.$$



Although there is no treatment effect, this will be non-zero if $n_T \neq n_P$. 
This bias is called *allocation bias* (see \@ref(sec:allocationbias)). Balance 
($n_T = n_P$) can only be guaranteed *on average across all possible trials*, 
but not for our specific trial. Hence, the trial design should ensure that 
$n_T \approx n_P$.


Stratification aims to control the imbalance between groups not with respect to
their size, but with respect to their *composition*. The idea is to use different
RPBs for different strata defined by relevant prognostic factors. For example,
if age is an important prognostic factor, one may use an RPB design for 
children and another one for adults. The number of children (and of adults)
receiving each treatment will then be very similar.

### Minimization

If there are many prognostic factors, using a separate RPB for each possible 
combination of these factors becomes impractical. For example, if there are 
5 binary prognostic factors, there will be already $2^5=32$ strata. 
*Minimization* aims to balance the groups with respect to each factor, but not for each combination of the factors. The minimization algorithm is:

-  Suppose a new patient enters the trial with certain values 
$x_1, \ldots, x_J$ of relevant prognostic factors.
-  The difference $D_j$ of numbers of patients allocated to group $A$ and patients allocated to group $B$ is computed for the observed value $x_j$ of each prognostic factor $j=1,\ldots,J$.
-  Compute the *total difference* $D = \sum_j D_j$ and proceed as follows (with some $p>1/2$): 
\[
\mbox{If} \left\{ \begin{array}{c} D = 0 \\ D < 0 \\ D > 0 \end{array} \right\}
\begin{array}{c}
  \mbox{then allocate the new patient} \\ \mbox{to group A with probability}
\end{array}
\left\{ \begin{array}{c} 0.5 \\ p \\ 1-p \end{array} \right\}.
\]



## Blinding and placebos

### Single and double blindness
Blinding is used to avoid assessment bias. We distinguish


-  *Single-blind trial*: patient is unware of the treatment being given.
-  *Double-blind trial*: neither doctor nor patient knows what treatment is being given.


In some trials the *statistical analysis* is also carried out *blinded*, because of subjective elements in any statistical analysis, e.g.

-  Should I transform the outcome variable? 
-  Should I compute risk differences, risk ratios or odds ratios for binary outcomes?
-  Should I include a covariate in the analysis?



### Placebos
Placebos are treatments that look similar to the true treatment but contain no active ingredient. They have essentially two roles: 


1. Placebos *facilitate blindness* in RCTs: A *treatment* versus *no treatment at all* comparison cannot be blinded, but  *treatment* versus *placebo* can often be blinded.

2.   Placebos help to control the *placebo effect*.


The *placebo effect* occurs if a patient exhibits a response to treatment, 
even though the treatment has *no active component* and cannot be having a 
direct effect. If only *untreated controls* (no treatment at all) are used, 
then we would not be able to tell whether an observed effect is due to treatment 
or to the placebo effect. Therefore a *placebo control group* is usually selected 
to control the placebo effect. In some trials, both untreated and placebo
control groups are used in addition to the usual intervention group to 
*directly assess* the size of the placebo effect.


### The double-dummy technique

Placebos can also help in the comparison of *two active treatments*. Suppose a 
blue tablet is to be compared to a red tablet. Color cannot be changed, but the
drug companies can produce placebo versions of the blue and red tablet, 
respectively. The *double-dummy technique* then gives both a blue *and* 
red tablet to each patient, in one group with blue active treatment and
red placebo and vice versa in the other group. For example, this method 
is useful to blind a comparison of a treatment given as a tablet with an 
intravenous treatment.



## Additional references
 
The design of experiments is discussed in @bland (Chapter 2). 
Allocation methods, assessment, blinding and placebos are discussed in 
@matthews (Chapters 4--5). Studies where the methods from this 
chapter are used in practice
are for example @berlin, @aardweg, @ballard.
