# Crossover trials 

Up to now, we have considered *parallel group designs*. 
A *crossover trial* trial compares the outcome of a patient when given 
treatment A to the outcome from the *same patient* when given treatment B. 
This means that participants act as their own controls, which leads to
larger precision of the treatment effect.
Crossover trials are possible if the aim of therapy is *not* to cure a condition. 
A typical application of the crossover design is to compare different
pain-relieving drugs. The simplest crossover trial design is the **AB/BA design**.


## The AB/BA design

The AB/BA design is illustrated in Table \@ref(tab:ABBA).
There are two treatment periods, and patients are randomized into two
sequence groups where patients in group AB receive the treatments in the
order A-B, and patients in group BA receive the treatments in the order B-A.

This design ensures that *period effects* can be separated from
*treatment effects*, which would not be the case if all patients
would have been allocated to group AB. For example, to study the
effect of pain-relieving drugs for headaches, a possible period effect
may be due to changes in weather conditions in the two periods.  If
all patients would be treated in the order A-B, then any difference in
the outcome between A and B may be due to the period effect alone, or
a combination of the period and treatment effect.



```{r abba-table, echo=FALSE}
library(knitr)

# Creating the data frame for the table
abba_data <- data.frame(
  Group = c("AB", "BA"),
  `Period 1` = c("A", "B"),
  `Period 2` = c("B", "A"), 
  check.names = FALSE
)

# Printing the table using kable
kable(abba_data, caption = "AB/BA design.", booktabs = TRUE, label = "ABBA")
```


Analysis methods for crossover trials with an AB/BA design are explained in the next sections, first for continuous outcomes and then for binary outcomes.


## Continuous outcomes

The following illustrating example will be used throughout this section.

```{r echo=F}
enuresis <- read.table("data/enuresis.csv", header=TRUE)
enuresis$diff <- enuresis$outcome1 - enuresis$outcome2
enuresis$group <- ifelse(enuresis$group==1, "AB", "BA")
enuresis$treatment <- ifelse(enuresis$group=="AB", enuresis$outcome1,  enuresis$outcome2)
enuresis$placebo <- ifelse(enuresis$group=="AB", enuresis$outcome2, enuresis$outcome1)
n <- nrow(enuresis)
n1 <- sum(enuresis$group=="AB")
n2 <- sum(enuresis$group=="BA")
``` 

:::{.example #enuresisex}
The Enuresis Trial [@matthews] is a placebo-controlled trial on $n=`r n`$ 
children suffering from enuresis (bed wetting). In group AB ($n_{AB}=`r n1`$), 
the drug is given for 14 days and the outcome (number of dry nights) is recorded.
Then, a placebo is administered for a fortnight and the same outcome variable is
recorded. In group BA ($n_{BA}=`r n2`$), first placebo and then the drug is given.
The outcome is treated as a continuous variable, although strictly speaking it
is a count variable with values between 0 and 14.
:::

```{r echo=TRUE}
head(enuresis, n=3)
``` 

Patient-level data for group AB and group BA comparing the outcomes under treatment 
and under placebo for each patient are shown in Figure \@ref(fig:enuresis)
together with the corresponding means.


```{r enuresis, fig.cap = "Patient-level data from the Enuresis Trial.", echo=F}
means <- lapply(split(enuresis, enuresis$group), 
       function(x) c(Mean = colMeans(x[,3:4])))

par(mfrow=c(1,2))
v3 <- enuresis$outcome1[enuresis$group=="AB"]
v4 <- enuresis$outcome2[enuresis$group=="AB"]
plot(0, 0, type = 'n', xlim = c(0.5, 2.5), ylim = c(0, 14), xaxt = 'n', xlab = "", ylab = "Number of dry nights", las = 1)
axis(1, at = c(1, 2), labels = c("Treatment", "Placebo"))
points(rep(1, length(v3)), v3, col = 2, pch = 16)
points(rep(2, length(v4)), v4, col = 2, pch = 16)
segments(rep(1, length(v3)), v3, rep(2, length(v4)), v4, col = 2, lwd = 2)
eps <- 0.15
segments(x0=1-eps, x1=1+eps, y0=means$A[1], y1=means$A[1], col=4, lwd=3)
segments(x0=2-eps, x1=2+eps, y0=means$A[2], y1=means$A[2], col=4, lwd=3)
text(1-eps, means$A[1], as.character(round(means$A[1],1)), col=4, pos=2)
text(2+eps, means$A[2], as.character(round(means$A[2],1)), col=4, pos=4)
title("Group AB")

v3 <- enuresis$outcome1[enuresis$group=="BA"]
v4 <- enuresis$outcome2[enuresis$group=="BA"]
plot(0, 0, type = 'n', xlim = c(0.5, 2.5), ylim = c(0, 14), xaxt = 'n', xlab = "", ylab = "Number of dry nights", las = 1)
axis(1, at = c(1, 2), labels = c("Placebo", "Treatment"))
points(rep(1, length(v3)), v3, col = 2, pch = 16)
points(rep(2, length(v4)), v4, col = 2, pch = 16)
segments(rep(1, length(v3)), v3, rep(2, length(v4)), v4, col = 2, lwd = 2)
segments(x0=1-eps, x1=1+eps, y0=means$B[1], y1=means$B[1], col=4, lwd=3)
segments(x0=2-eps, x1=2+eps, y0=means$B[2], y1=means$B[2], col=4, lwd=3)
text(1-eps, means$B[1], as.character(round(means$B[1],1)), col=4, pos=2)
text(2+eps, means$B[2], as.character(round(means$B[2],1)), col=4, pos=4)
title("Group BA")
legend("bottomright", col=4, lty=1,  legend="means", lwd=3, bty="n")
``` 


### Simple analysis {#sec:simple}

First, we set the notation as follows:

-  $x_{ij}$ is the (continuous) outcome of patient $i$ in period $j$, for $i=1,\ldots, n$,
-  $\alpha$ is the mean outcome in period 1 under placebo (treatment B),
-  $\Delta$ is the treatment effect of the drug relative to placebo, 
-  $\beta$ is the period effect of period 2 relative to period 1. 

The mean outcomes are therefore modeled as:

```{r, echo = FALSE}
# Creating the data frame for the table
period_data <- data.frame(
  Period = c(1, 2),
  `Group AB` = c("$\\E(x_{i1}) = \\alpha + \\Delta$", "$\\E(x_{i2}) = \\alpha + \\beta$"),
  `Group BA` = c("$\\E(x_{i1}) = \\alpha$", "$\\E(x_{i2}) = \\alpha + \\Delta + \\beta$"), 
  check.names = FALSE
)

# Printing the table using kable
kable(period_data, escape = FALSE, booktabs = TRUE)
```


Now consider the *within-patient differences* $d_i = x_{i1} - x_{i2}$ and let $\bar d_{AB}$ and $\bar d_{BA}$ denote the mean difference in group AB and group BA, respectively. Then:

\begin{eqnarray*}
\E(\bar d_{AB}) & = & \, \, \, \, \Delta - \beta \mbox{ in group AB}\\
\mbox{ and } \E(\bar d_{BA}) & = & - \Delta - \beta \mbox{ in group BA},  \\
\mbox{ so } \E(\bar d_{AB} -\bar d_{BA} )& = & 2 \Delta \, . 
\end{eqnarray*}

The null hypothesis of no treatment difference, $\Delta=0$, can hence be investigated with a two-sample $t$-test applied to the two sets of within-patient differences. The treatment effect $\Delta$ is finally estimated as half of the differences in means: $\hat \Delta = (\bar d_{AB} -\bar d_{BA})/2$.

:::{.example #enuresisex name="continued"}
Results of the simple analysis in the Enuresis Trial:
:::


```{r echo=TRUE}
(simpleAnalysis <- t.test(diff ~ group, data = enuresis, var.equal=TRUE))
## treatment effect: divide by 2
(Delta <- mean(simpleAnalysis$conf.int)/2)  
## confidence interval for treatment effect: divide by 2
(DeltaConfInt <- simpleAnalysis$conf.int/2)
```


Just for illustration, but *not recommended*: 
A naive analysis of the AB/BA design would be to compare differences $d_i$ of 
treatment to placebo measurements with a paired $t$-test, ignoring group membership:
    \[
    d_i = \left\{ \begin{array}{rl}
      x_{i1} - x_{i2} & \mbox{ in group AB} \\
      x_{i2} - x_{i1} & \mbox{ in group BA} \\
      \end{array}\right.
      \]
The mean $\bar d$ then has expectation
\begin{eqnarray*}
  \E(\bar d) &=& \frac{1}{n_{AB}+n_{BA}} \left\{n_{AB} (\Delta - \beta) + n_{BA} (\Delta + \beta) \right\}
  \\ & = & \Delta - \beta \, \frac{n_{AB}-n_{BA}}{n_{AB}+n_{BA}}  
\end{eqnarray*}

The estimate from the t-test is confounded by the period effect $\beta$.
Bias occurs whenever group sizes are unequal ($n_{AB} \neq n_{BA}$) and there is 
a period effect ($\beta \neq 0$).

:::{.example #enuresisex name="continued"}
Results of the non-recommended paired $t$-test in the Enuresis Trialb:
:::

```{r echo=TRUE}
(naiveAnalysis <- t.test(enuresis$treatment, enuresis$placebo, paired=TRUE))
(Delta.conf.int <- naiveAnalysis$conf.int)
```

The paired $t$-test gives different results and is **not recommended**.


#### Between-patient variation {-}

The previously described (recommended) analysis also holds if we allow for *patient-specific effects* $\xi_i$:

```{r, echo = FALSE}
# Creating the data frame for the table
period_data <- data.frame(
  Period = c(1, 2),
  `Group AB` = c("$\\E(x_{i1})  =  \\alpha + \\Delta + {\\color{red}\\xi_i}$", 
                 "$\\E(x_{i2})  =  \\alpha + \\beta + {\\color{red}\\xi_i}$"),
  `Group BA` = c("$\\E(x_{i1})  =  \\alpha + {\\color{red}\\xi_i}$", 
                 "$\\E(x_{i2})  =  \\alpha + \\Delta + \\beta + {\\color{red}\\xi_i}$"), 
  check.names = FALSE
)

# Printing the table using kable
kable(period_data, escape = FALSE, booktabs = TRUE)
```


since the $\xi_i$'s cancel when we calculate patient-specific differences 
$d_{i} = x_{i1} - x_{i2}$. This illustrates that between-patient variation is eliminated in the standard analysis of crossover trials.

### Analysis using mixed models

It is also possible to perform an analysis of the original outcomes $x_{ij}$ using a *mixed model* with *patient-specific random effects* $\xi_i$. This will give the same treatment effect (with confidence interval and $p$-value) as the simple analysis based on the $t$-test, but also provides an estimate of the period effect and estimates of within- and between-patient variances. 


:::{.example #enuresisex name="continued"}
Results in the Enuresis Trial:
:::

```{r echo=TRUE, warning=FALSE}
# restructuring the data
outcome <- c(enuresis$outcome1, enuresis$outcome2)
n <- nrow(enuresis)
period <- as.factor(c(rep(1, n), rep(2, n)))
id <- c(enuresis$id,enuresis$id)
treatment <- as.numeric(c((enuresis$group=="AB"),
                          (enuresis$group=="BA")))

## fit mixed model in R
library(lme4)
mixed1 <- lmer(outcome ~ period + treatment + (1|id))
print(coef(summary(mixed1)))

## comparison to standard analysis
print(mean(simpleAnalysis$conf.int/2))
print(simpleAnalysis$stderr/2)
``` 

The mixed model gives the same estimate of the treatment effect as the simple analysis based on the comparison of patient-specific differences with a  $t$-test, but
also provides an estimate of the period effect ($\hat \beta = `r coef(summary(mixed1))[2,1]`$).

The approach also gives estimates of the variance (respectively standard deviation) components: 
```{r echo=TRUE}
summary(mixed1)$varcor
``` 

The estimated between-patient standard deviation $\sigma_b = `r round(sqrt(summary(mixed1)$varcor[[1]]),2)`$ is denoted by \code{id} and the estimated
within-patient standard deviation $\sigma_w = `r round((summary(mixed1)$sigma),2)`$ is denoted by Residual.


## The issue of carryover


The above analysis assumes that there is no *carryover effect*, i.e.
effects of the treatment given in period 1 do not persist during period 2.
We now describe a statistical approacha to test for carryover 
in a AB/BA crossover trial with continuous outcome. We emphasize that this is generally not recommended due to lack of power and other problems. 
The recommended approach is not to use a crossover design when there is a 
possibility of a carryover effect. 
You should try to use non-statistical arguments, perhaps based on the half-lives of drugs, etc., to decide how long treatment effects are likely to persist and
apply appropriate washout periods between the two treatment periods. 



### Analysis of carryover

Suppose now we have an additional carryover effect $\gamma$ in period 2 for patients in group AB, but not in group BA:

```{r, echo = FALSE}
# Creating the data frame for the table
period_data <- data.frame(
  Period = c(1, 2),
  `Group AB` = c("$\\E(x_{i1})  =  \\alpha + \\Delta$", 
                 "$\\E(x_{i2})  =  \\alpha + \\beta {\\color{red} \\,+\\, \\gamma}$"),
  `Group BA` = c("$\\E(x_{i1})  =  \\alpha$", 
                 "$\\E(x_{i2})  =  \\alpha + \\Delta + \\beta$")
)

# Printing the table using kable
kable(period_data, escape = FALSE, booktabs = TRUE, label = NA) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```

Then:

\begin{eqnarray*}
\E(\bar d_{AB}) & = & \, \, \, \, \Delta - \beta {\color{red} \, - \, \gamma} \mbox{ in group AB},\\
\E(\bar d_{BA}) & = & - \Delta - \beta \mbox{ in group BA},  \\
\mbox{ so } \E(\bar d_{AB} -\bar d_{BA} )& = & 2 \Delta {\color{red} \, - \, \gamma} \, .
\end{eqnarray*}

So, for $\gamma \neq 0$, the traditional estimate of $\Delta$ will be biased.


### Test for carryover

A test for $H_0$: $\gamma = 0$ can be performed using a standard $t$-test comparing the sums $s_i=x_{i1} + x_{i2}$ across groups:

\begin{eqnarray*}
\E(\bar s_{AB}) & = & 2 \alpha + \Delta + \beta + \gamma \mbox{ in group AB},\\
\E(\bar s_{BA}) & = & 2 \alpha + \Delta + \beta \mbox{ in group BA},  \\
\mbox{ so } \E(\bar s_{AB} - \bar s_{BA} )& = & \gamma 
\end{eqnarray*}


As already mentioned, this procedure is **not recommended**. Instead, non-statistical arguments should be used to decide how long treatment effects are likely to persist.

:::{.example #enuresisex name="continued"}
Standard test for carryover in the Enuresis Trial:
:::

 
```{r echo=TRUE}
enuresis$sum <- enuresis$outcome1 + enuresis$outcome2
(res <- t.test(sum ~ group, data = enuresis, var.equal=TRUE))
(gamma <- mean(res$conf.int))
(gamma.conf.int <- res$conf.int)
``` 

Analysis using mixed models gives identical estimates of the carryover effect and the same value for the $t$-statistic:

```{r echo=TRUE}
carryover <- ifelse((period==2) & (treatment==0), 1, 0)
res3 <- lmer(outcome ~ period + treatment + carryover + (1|id))
print(coef(summary(res3)))
``` 


```{r echo=FALSE}
c <- coef(summary(res3))[,1]
``` 

This is as so-called *saturated model*, as it fits the four
parameters $\alpha$ (\code{Intercept}), $\beta$ (\code{period}), $\Delta$ (\code{treatment}), and
$\gamma$ (\code{carryover}) to the
four data entries (the means in each cell).  The fitted values
in the four cells are therefore equal to the means shown in Figure \@ref(fig:enuresis) (up to rounding errors):


```{r}
# Creating the data frame for the table
period_data <- data.frame(
  Period = c(1, 2),
  `Group AB` = c(
    sprintf("$\\E(x_{i1})  =  %.3f = %.3f + %.3f$", round(means$A[1], 3), round(c[1], 3), round(c[3], 3)),
    sprintf("$\\E(x_{i2})  =  %.3f = %.3f + %.3f + %.3f$", round(means$A[2], 3), round(c[1], 3), round(c[2], 3), round(c[4], 3))
  ),
  `Group BA` = c(
    sprintf("$\\E(x_{i1})  =  %.3f$", round(means$B[1], 3)),
    sprintf("$\\E(x_{i2})  =  %.3f = %.3f + %.3f + %.3f$", round(means$B[2], 3), round(c[1], 3), round(c[3], 3), round(c[2], 3))
  )
)

# Printing the table using kable
kable(period_data, escape = FALSE, booktabs = TRUE, label = NA) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```


Interpretation of the coefficients of a model with carry-over effect is difficult. The treatment effect is now assumed to interact with period, so is no longer constant across the two periods. 
The negative sign of the carryover effect is a consequence of a systematic difference in the outcome means between the AB and the BA group. For a perfectly randomised trial we would expect the sum of the two outcomes across the two treatment to be the same, but in the enuresis trial they are
`r round(2*mean(means$A),1)` dry days in the AB group and `r round(2*mean(means$B),1)` days in the BA group, and the difference of the two is the carryover effect of 
`r round(2*mean(means$A)-2*mean(means$B),1)` days. 




## Sample size for AB/BA design

We now discuss sample size calculation for a AB/BA crossover design
with continuous outcome. Consider the model with patient-specific
random-effects $\xi_i$:
  \[
    x_{ij} = \mbox{ fixed effects } + \xi_i + \epsilon_{ij}
  \]
We distinguish

-  the **between-patient** variance $\sigma_b^2 = \Var(\xi_i)$ and
-  the **within-patient** variance $\sigma_w^2 = \Var(\epsilon_{ij})$. 


The variance of the within-patient differences $d_i = x_{i1} - x_{i2}$ is $\sigma_d^2=2\sigma_w^2$ and we want to detect $\Delta_d = 2 \Delta$ with a standard hypothesis test. The total variance is $\sigma_{tot}^2 = \sigma_b^2 + \sigma_w^2 = \sigma_w^2 /(1-\rho)$ with intraclass correlation coefficient 
$\rho=\sigma_b^2/(\sigma_b^2 + \sigma_w^2)$. The sample size per group for a crossover trial is hence
    \[
    n_{\tiny{\mbox CO}} = \frac{2 \sigma_d^2 (u + v)^2}{\Delta_d^2} = \frac{\sigma_w^2 (u + v)^2}{ \Delta^2}. 
    \]

#### Comparison with parallel group design {-}
 
The **total variance** of $x_{ij}$ is
  \[
   \sigma^2 = \Var(x_{ij}) = \sigma_b^2 + \sigma_w^2 = \sigma_w^2 /(1-\rho)
  \]
with intraclass correlation coefficient 
$\rho=\sigma_b^2/(\sigma_b^2 + \sigma_w^2)$.
The standard parallel group design thus requires
  \[
    n_{\tiny{\mbox PG}} = \frac{2 \sigma^2 (u + v)^2}{ \Delta^2} = \frac{2 \sigma_w^2 (u + v)^2}{(1-\rho) \Delta^2}
    \]
patients per group. Less patients are needed in a crossover trial, 

$$n_{\tiny{\mbox CO}}/n_{\tiny{\mbox PG}} = (1-\rho)/2,$$

but note that *two measurements* per patient are required in the crossover trial compared to only one in the parallel group design.


## Binary outcomes

Consider the following illustrating example that will be used throughout this section:

:::{.example #cerebro}
In a $2 \times 2$ crossover trial on cerebrovascular deficiency with 67 patients, an active treatment is compared to placebo [@Jones2014]. The outcome is whether an electrocardiogram was judged normal or abnormal. 
:::

```{r echo=FALSE}
## install.packages("teachingUZH_1.0-5.tar.gz", repos = NULL)
library(teachingUZH)
data(cerebrovascular)
## for didactical reasons
## change response for ID=22 to avoid zero counts
cerebrovascular[43,]$y <- 0

cerebrovascular$treatment <- factor(cerebrovascular$treatment, levels = 0:1, labels = c("Placebo", "Active"))
cerebrovascular$y.f         <- factor(cerebrovascular$y, levels = 0:1, labels = c("Abnormal", "Normal"))

``` 

```{r echo=TRUE}
head(cerebrovascular)
tail(cerebrovascular)
``` 


For illustration purposes we start with a separate analysis by period. 
In period 1, each patient receives either the active treatment or the
placebo and the same holds for period 2. So, by considering the two
periods separately, all observations are independent as we are used to
from a parallel group design. 

```{r echo = F}
library(xtable)

tab1 <- table("treatment"=cerebrovascular$treatment[cerebrovascular$time==0], 
              "Period 1"=cerebrovascular$y.f[cerebrovascular$time==0])
tab2 <- table(cerebrovascular$treatment[cerebrovascular$time==1], 
              cerebrovascular$y.f[cerebrovascular$time==1])


pair <- group <- character()
cere <- cerebrovascular

for(i in 1:max(cere$ID)){
    y1 <- cere$y.f[i*2-1]
    y2 <- cere$y.f[i*2]
    if((y1=="Normal")&(y2=="Normal")) pair[i] <- "No-No"
    if((y1=="Normal")&(y2=="Abnormal")) pair[i] <- "No-Ab"
    if((y1=="Abnormal")&(y2=="Normal")) pair[i] <- "Ab-No"
    if((y1=="Abnormal")&(y2=="Abnormal")) pair[i] <- "Ab-Ab"
    group[i] <- ifelse(cere$treatment[i*2-1]=="Active", "AB", "BA")
    }
fullTable <- table(group, pair)[,c(4:1)]

est <- fisher.test(tab1)$estimate
se <- sqrt(sum(1 / tab1))

est2 <- fisher.test(tab2)$estimate
se2 <- sqrt(sum(1 / tab2))
``` 

```{r echo=F}
library(biostatUZH)
library(Epi)
```

Period 1:

```{r period1, results='asis', echo=FALSE}
library(xtable)

kable(tab1, booktabs = TRUE, caption = "Period 1", align = "c", 
      label = NA) %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)

```

\begin{eqnarray*}
    \text{OR}_1 &=& \frac{`r tab1[1, 1]` \cdot `r tab1[2, 2]`}{`r tab1[1, 2]` \cdot `r tab1[2, 1]`} \ = \ `r round(est, 2)`\\
    \text{se(log(OR$_1$))} &=& \sqrt{\tfrac{1}{`r tab1[1, 1]`} + \tfrac{1}{`r tab1[1, 2]`} + \tfrac{1}{`r tab1[2, 1]`} + \tfrac{1}{`r tab1[2, 2]`}} = `r round(se, 2)`\\
    \text{95% CI} &=& `r paste("[",  round(exp(log(est) - qnorm(0.975) * se), 2), ", ",  round(exp(log(est) + qnorm(0.975) * se), 2), "]", sep = "")`
\end{eqnarray*}

Period 2:

```{r}
# Print the contingency table using xtable
kable(tab2, booktabs = TRUE, caption = "Period 2", align = "c", 
      label = NA) %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)

```

\begin{eqnarray*}
    \text{OR}_2 &=& \frac{`r tab2[1, 1]` \cdot `r tab2[2, 2]`}{`r tab2[1, 2]` 
    \cdot `r tab2[2, 1]`} \ = \ `r round(est2, 2)` \\
    \text{se(log(OR$_2$))} &=& \sqrt{\tfrac{1}{`r tab2[1, 1]`} + 
    \tfrac{1}{`r tab2[1, 2]`} + \tfrac{1}{`r tab2[2, 1]`} +
    \tfrac{1}{`r tab2[2, 2]`}} = `r round(se2, 2)`\\
    \text{95% CI} &=& `r paste("[", round(exp(log(est2) - qnorm(0.975) * se2), 2), ", ", round(exp(log(est2) + qnorm(0.975) * se2), 2), "]", sep = "")`
\end{eqnarray*}



The confidence intervals for the odds ratios in the two analyses are rather 
wide, providing no evidence for a treatment effect in the two periods. 
A *combined analysis* has to take into account that responses from the 
same patient are correlated.
The results of a crossover trial with binary outcome can be summarized as in 
the following table:

```{r }
sel1 <- which((cerebrovascular$treatment=="Active" & cerebrovascular$time==0) | (cerebrovascular$treatment=="Placebo" & cerebrovascular$time==1))
sel2 <- which((cerebrovascular$treatment=="Placebo" & cerebrovascular$time==0) | (cerebrovascular$treatment=="Active" & cerebrovascular$time==1))

group1 <- cerebrovascular[sel1,]
group2 <- cerebrovascular[sel2,]

tab1 <- table("period"=group1$time, "response"=group1$y.f)
tab2 <- table("period"=group2$time, "response"=group2$y.f)
``` 

```{r}
library(kableExtra)
# Define variables
a <- fullTable[1, 2]
b <- fullTable[1, 3]
c <- fullTable[2, 2]
d <- fullTable[2, 3]

# Create the table
kable(
  data.frame(
    Group = c("Active-Placebo", "Placebo-Active"),
    `No-No` = c(fullTable[1, 1], fullTable[2, 1]),
    `No-Ab` = c(paste0("**a =** ", a), paste0("**c =** ", c)),
    `Ab-No` = c(paste0("**b =** ", b), paste0("**d =** ", d)),
    `Ab-Ab` = c(fullTable[1, 4], fullTable[2, 4])
  ),
  col.names = c("Group", "No-No", "No-Ab", "Ab-No", "Ab-Ab"),
  format = "html", 
) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
add_header_above(c(" " = 1, "Outcome pair" = 4))

# # Print table using kable
# knitr::kable(fullTable, booktabs = TRUE) %>%
#   kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
```



Only discordant pairs (in bold) contribute to estimates of the
treatment effect. Group imbalance occurs if $a + b \neq c + d$.

<!-- A combined analysis can be done using Mainland-Gart test  -->
<!-- or a generalized linear mixed model. -->


### Naive analysis

A naive way to analyse such data would be to compare the treatment 
groups using Mc Neymar's test for binary paired data, ignoring the 
group membership. This approach assumes that there is 
no period effect nor group imbalance.\todo{CM: in slides: balance?}
The estimate of the odds ratio OR (Active vs. Placebo) 
is based on the number of discordant pairs in the expected 
vs. unexpected direction 

\begin{eqnarray*} 

\widehat{\OR} = \frac{\mbox{ \# pairs: normal for Active, abnormal for Placebo}}{\mbox{ \# pairs: abnormal for Active, normal for Placebo}} = \frac{a + d}{b + c}

\end{eqnarray*}

with standard error

\begin{eqnarray*}
\SE(\log \widehat{\OR}) = \sqrt{\frac{1}{a+d} + \frac{1}{b+c}}.
\end{eqnarray*}

```{r echo=FALSE}
subTable <- fullTable[,c("No-Ab", "Ab-No")]
a <- subTable[1,1]
b <- subTable[1,2]
c <- subTable[2,1]
d <- subTable[2,2]
``` 

```{r echo=TRUE}

or <- (a+d)/(b+c)
se.log.or <- sqrt(1/(a+d) + 1/(b+c))
printWaldCI(log(or), se.log.or, FUN=exp, digits=2)

``` 

It is also possible to perform Mc Neymar test using the `R` function 
`mcnemar.test()`: 
```{r echo=FALSE}

x <- as.table(matrix(c(39, 10, 3, 15), ncol=2, byrow=TRUE), dnn=)
rownames(x) <- colnames(x) <- c("Normal", "Abnormal") 

``` 



```{r echo=TRUE}

print(x)
mcnemar.test(x)
mcnemar.test(x, correct=FALSE)

``` 

However, no effect estimate is given, and the test is based on a
slighlty different test statistic (with or without continuity correction).


### Recommended analysis using Mainland-Gart test

To incorporate a possible period effect, 
we compare discordant pairs in each sequence group: 

\begin{eqnarray*} 

\widehat{\OR} = \left({\frac{\OR \mbox{ in group Active-Placebo}}{\OR \mbox{ in group Placebo-Active}}}\right)^{1/2} = \left({\frac{a/b}{c/d}}\right)^{1/2}

\end{eqnarray*} 


with standard error

$$\SE( \log \widehat{\OR}) = \frac{1}{2} \sqrt{\frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}}.$$

Note the additional factor $1/2$ in both $\widehat{\OR}$ and $\SE(\hat \beta)$, as in the analysis of a continuous outcome. 


```{r echo=TRUE}
or <- sqrt((a/b)/(c/d))
se.log.or <- 1/2 * sqrt(1/a + 1/b + 1/c + 1/d)
printWaldCI(log(or), se.log.or, FUN=exp, digits=2)

``` 
  
```{r echo=FALSE, eval = TRUE}
# print(fullTable)
# print(subTable <- fullTable[,c(2,3)])

a <- subTable[1,1]
b <- subTable[1,2]
c <- subTable[2,1]
d <- subTable[2,2]

log.or <- log(a*d/(b*c))/2
se.log.or <- sqrt(1/a+1/b+1/c+1/d)/2
# printWaldCI(log.or, se.log.or, FUN=exp, digits=3)
``` 



### Analysis with generalized linear mixed models
We may also use a generalized linear mixed model (but results may depend on the choice of the integration parameter \code{nAGQ}):

```{r echo=TRUE}
resCerebro <- glmer(y ~ treatment + time + (1|ID), 
                     family=binomial, data=cerebrovascular, nAGQ=10)

(glmmTable <- coef(summary(resCerebro)))

printWaldCI(glmmTable[2,1], glmmTable[2,2], FUN=exp, digits=2)
``` 

```{r echo = TRUE}

resCerebro <- glmer(y ~ treatment + time + (1|ID), 
                     family=binomial, data=cerebrovascular, nAGQ=5)

(glmmTable <- coef(summary(resCerebro)))

printWaldCI(glmmTable[2,1], glmmTable[2,2], FUN=exp, digits=2)

```


## Additional references

You can find more about crossover trials in @bland  (Ch. 2.7) and in @matthews 
(Ch. 11).
Practical examples of crossover trials discussed in the lab are 
@frank2008 and @allan2001. More details on crossover trials are given in 
@Senn as well as in @Senn2 (Chapter 17). 



