# Meta-analysis 
Results from a single study are not sufficient to detect or exclude a
relevant effect of a new therapy. Meta-analysis describes statistical
techniques to summarize the effect estimates obtained from several
independent studies on the same research question [@Schmid_etal2021].
The studies
included in a meta-analysis are usually the result of a systematic
review [@Egger_etal2022]. There are also narrative systematic reviews where
a subsequent meta-analysis is not possible. 


## Systematic review

A *systematic review* aims to identify relevant studies from a
literature review.  This is usually done with a literature search with
pre-defined search terms in a database such as *MEDLINE*. Results from
unpublished studies may also be included, if they are available from platforms
such as *clinicaltrials.gov*. The goal of a subsequent
meta-analysis is to summarize the results from a number of clinical
studies on the same research question. A
systematic review with subsequent meta-analysis proceeds as follows:

1. Systematically review the available evidence.
2. Provide quantitative summaries of the results from each study.
3. Combine these results across studies, if appropriate (meta-analysis).
4. Provide summary effect estimate. 

This is the basis of evidence-based medicine (EBM).
The studies included in a systematic review should be sufficiently
homogeneous regarding in- and exclusion criteria and should use the
same effect measure. A typical effect measure for continuous
outcomes is the (standardized)
difference in means $\theta$ between treatment groups.
For binary/time to event outcomes, relative treatment
effects are preferred, such as relative risk $\RR$, odds ratio $\OR$
and hazard ratio $\HR$. These are usually considered on a log-scale,
so the relevant effect estimate is $\theta=\log(\RR)$,
$\theta=\log(\OR)$, $\theta=\log(\HR)$.

:::{.example #preeclampsia}
Consider nine RCTs studying the incidence of preeclampsia in 
patients treated with diuretic versus a control treatment.
The reported effect measure is an odds ratio (diuretic vs. control), 
see Table \@ref(tab:preeclampsia).
:::

```{r preeclampsia, echo=FALSE}
library(knitr)
kable(
  data.frame(
    Study = c("Weseley", "Flowers", "Menzies", "Fallis", "Cuadros", "Landesman", "Krans", "Tervila", "Campbell"),
    Diuretic = c("11% (14/131)", "5% (21/385)", "25% (14/57)", "16% (6/38)", "1% (12/1011)", "10% (138/1370)", "3% (15/506)", "6% (6/108)", "42% (65/153)"),
    Control = c("10% (14/136)", "13% (17/134)", "50% (24/48)", "45% (18/40)", "5% (35/760)", "13% (175/1336)", "4% (20/524)", "2% (2/103)", "39% (40/102)"),
    OR = c(1.04, 0.40, 0.33, 0.23, 0.25, 0.74, 0.77, 2.97, 1.14)
  ),
  caption = "Results of nine RCTs about incidence of preeclampsia comparing diuretic vs. control.",
  col.names = c("Study", "Diuretic", "Control", "OR"),
  booktabs = TRUE, align = "l"
)
```



```{r ch1, echo=F}
preeclampsia <- read.table("data/preeclampsia.txt", header = TRUE)

diuretics <- with (preeclampsia,
                   c (sum (Diuretic[Preeclampsia == "yes"]), sum (Diuretic[Preeclampsia == "no"]))
                   )
controls <- with (preeclampsia,
                   c (sum (Control[Preeclampsia == "yes"]), sum (Control[Preeclampsia == "no"]))
                   )
logOddsRatio <- log (diuretics[1] * controls[2] / diuretics[2] / controls[1])
standardError <- sqrt (sum (1/diuretics) + sum (1/controls))
waldKiLogOddsRatio <- logOddsRatio + c (-1, 1) * 1.96 * standardError
waldData <- list (logOddsRatio, waldKiLogOddsRatio)

## compute data
oddsRatio <- function (square)
    (square[1,1] * square[2,2]) / (square[1,2] * square[2,1])

variance <- function (square)
    sum (1 / square)

groups <- split (subset (preeclampsia, select = c (Diuretic, Control)),
                 preeclampsia[["Trial"]])

logOddsRatios <- log (sapply (groups, oddsRatio))
variances <- sapply (groups, variance)
w <- 1/variances
num <- sum(w*logOddsRatios)
den <- sum(w)
theta.hat <- num/den
se.theta.hat <- 1/sqrt(sum(w))
wald.z <- (theta.hat)/se.theta.hat
wald.p <- 2*pnorm(abs(wald.z), lower.tail=FALSE)

t.homo <- sum(w*(logOddsRatios-theta.hat)^2)
n <- length(logOddsRatios)
p.homo <- pchisq(t.homo, df=n-1, lower.tail=FALSE)

``` 

Figure \@ref(fig:preeclampsiaFig) shows a graphical summary of the log odds ratios with 95\% CIs. This kind of graphical summary is called a *forest plot*.


```{r preeclampsiaFig, fig.cap = "Forest plot for the estimated effects with 95\\% confidence intervals in the nine studies from Example \\@ref(exm:preeclampsia).", echo=FALSE}


## plot intervals with point estimates
library (lattice)

## study effects
postSd <- sqrt (variances)
postLower <- logOddsRatios - postSd * 1.96
postUpper <- logOddsRatios + postSd * 1.96

## plot intervals with point estimates

panel.ci <- function(x, y, lx, ux, subscripts, ...)
{
    x <- as.numeric(x)
    y <- as.numeric(y)

    lx <- as.numeric(lx[subscripts])
    ux <- as.numeric(ux[subscripts])

    panel.dotplot(x, y, lty = 2, ...)            # normal dotplot for point estimates
    panel.arrows(lx, y, ux, y,          # draw intervals
                 length = 0.1, unit = "native",
                 angle = 90,            # deviation from line
                 code = 3,              # left and right whisker
                 ...)
    panel.abline (v = 0, lty = 2)       # reference line
}

studyNames <- c (names (postLower))
studyNames <- ordered (studyNames, levels = rev (studyNames)) # levels important for order!

ciData <- data.frame (low = postLower, 
                      up = postUpper,
                      mid = logOddsRatios,
                      names = studyNames
                      )
ciData[["signif"]] <- with (ciData,
                            up < 0 | low > 0)
ciData[["farbe"]] <- with (ciData,
                           ifelse (signif, "black", "black"))

randomEffectsCiPlot <- with (ciData,
                             dotplot (names ~ mid,
                                      panel = panel.ci,
                                      lx = low, ux = up,
                                      pch = 19, col = farbe,
                                      xlim = c (-3, 3), xlab = "Log Odds Ratio",
                                      scales = list (cex = 1)
                                      )
                             )
print (randomEffectsCiPlot)
```



## Meta-analysis

We set the following notation for a *meta-analysis* of $i=1,\ldots, n$ trials:

-  $\theta_i$, $\hat \theta_i$ are the true and estimated study-specific treatment effect,
-  $s_i = \se(\hat \theta_i)$ is the standard error of $\hat \theta_i$,
-  $v_i = s_i^2$ is the variance of $\hat \theta_i$ and

There are two different types of analyses: fixed and random effects model.

### Fixed effect model

A fixed effect model is based on the *homogeneity assumption*: 
$\theta_i = \theta$ for all $i$.
The estimate $\hat \theta$ of the overall treatment effect $\theta$ is
then a weighted average of study-specific estimates 
$\hat \theta_i$ with inverse variance weights $w_i=1/v_i$:

\begin{equation*}
  \hat \theta = \frac{\sum w_i \hat \theta_i}{\sum w_i}
\mbox{  with }
  \se(\hat \theta) =  1/\sqrt{\sum w_i}.
\end{equation*}


```{r ch4.0, echo = FALSE}
pree <- read.csv("data/preeclampsia.csv", header = TRUE, sep = ",", dec = ".", fill = TRUE)     
pree <- pree[order(pree$study),]

#transformVarNames(pree, "pree")
study       <- pree$study
treatedPre <- pree$diur_pre
treatedTot <- pree$diur_tot
controlPre <- pree$plac_pre
controlTot <- pree$plac_tot
a <- treatedPre
b <- treatedTot
c <- controlPre
d <- controlTot
``` 



The overall effect is usually displayed in the forest plot, as in
Figure \@ref(fig:preeclampsiaMA) for the preeclampsia
Example \@ref(exm:preeclampsia). This figure is generated with the following 
`R` code:
```{r ch4b.0, echo = TRUE, fig.show='hide'}
library(meta)
meta1 <- metabin(event.e = treatedPre, n.e = treatedTot, 
                 event.c = controlPre, n.c = controlTot, 
                 sm = "OR" , method = "Inverse", studlab = study)
forest(meta1, comb.fixed = TRUE, comb.random = TRUE) 
``` 


```{r preeclampsiaMA, fig.cap = "Results with forest plot from package meta for a fixed effect analysis in Example \\@ref(exm:preeclampsia)", echo = FALSE, fig.height=7.5, fig.width=9.75}
library(meta)
meta1 <- metabin(event.e = treatedPre, n.e = treatedTot, 
                 event.c = controlPre, n.c = controlTot, 
                 sm = "OR" , method = "Inverse", studlab = study)
forest(meta1, comb.fixed = TRUE, comb.random = TRUE) 
``` 


```{r ch5.0, echo=F}
or.fixed <- exp(meta1$TE.fixed)
or.lower <- exp(meta1$lower.fixed)
or.upper <- exp(meta1$upper.fixed)
ci.fixed <- c(or.lower, or.upper)

or.random <- exp(meta1$TE.random)
or.lower <- exp(meta1$lower.random)
or.upper <- exp(meta1$upper.random)
ci.random <- c(or.lower, or.upper)
``` 



#### Cochran's test for heterogeneity {-}
Under the homogeneity assumption, we have

\begin{equation*}
Q = \sum w_i (\hat \theta_i - \hat \theta)^2 \sima \chi^2_{n-1}
\end{equation*}

This can be used to calculate the $P$-value of 
*Cochran's test for heterogeneity*.

:::{.example #preeclampsia name="continued"}
For the preeclampsia data, this test yields 
$Q = `r round(t.homo, 1)`$ at $n-1=8$ degrees of freedom
($p=`r formatPval(p.homo,digits=1)`$). 
There is strong evidence for heterogeneity between studies, so
a fixed effect model is questionable.
:::


### Random effects model
A random effects model assumes that the $\theta_i$'s come from a normal
distribution with mean $\theta^*$ and *heterogeneity variance* $\tau^2$:

\begin{equation*}
\hat \theta_i \given \theta_i  \sim   \Nor(\theta_i, v_i) \quad \mbox{ and }
\quad \theta_i   \sim   \Nor(\theta^*, \tau^2),
\end{equation*}

so marginally

\begin{equation*}
\hat \theta_i \sim   
\Nor(\theta^*, v_i + \tau^2).
\end{equation*}

The overall effect estimate $\hat \theta^*$ and its standard error are now:

\begin{equation*}
  \hat \theta^*  = \frac{\sum w_i^* \hat \theta_i}{\sum w_i^*},
  \quad
  \se(\hat \theta^*) = {1}/{\sqrt{\sum w_i^*}},
\end{equation*}

with weights $w_i^*  =  {1}/{(v_i + \tau^2)}$. 
Compared to the fixed effect model, the CIs for the overall effect will become
wider and large studies will obtain less weight.


#### Estimate of heterogeneity variance and Higgins' $I^2$ {-}

The *moment estimator* of the heterogeneity variance compares the 
$Q$-statistic to its expectation $n-1$ under homogeneity.
Truncation at zero and appropriate scaling gives 

\begin{equation*}
  \hat \tau^2 = \left\{Q-(n-1)\right\}_{+} / \left\{\textstyle\sum w_i - \textstyle\sum w_i^2/\textstyle\sum w_i\right\}.
\end{equation*}

The standard **DerSimonian-Laird** approach plugs $\hat \tau^2$
in $w_i^* = {1}/{(v_i + \tau^2)}$, 
the alternative **Hartung-Knapp** adjustment accounts for
the uncertainty of $\hat \tau^2$. 


```{r ch3, echo=F}
Q <- t.homo
sigma2.hat <- max(0, (Q-(n-1))/(sum(w)-sum(w^2)/sum(w)))
w.star <- 1/(variances+sigma2.hat)
num.star <- sum(w.star*logOddsRatios)
den.star <- sum(w.star)
theta.star.hat <- num.star/den.star
se.theta.star.hat <- 1/sqrt(sum(w.star))
wald.star.z <- (theta.star.hat)/se.theta.star.hat
wald.star.p <- 2*pnorm(abs(wald.star.z), lower.tail=FALSE)
``` 

Easier to interpret is *Higgins' $I^2$*, the percentage of variance that 
is attributable to study heterogeneity.  The value of $I^2$ will somewhat 
depend on the method used to estimate $\tau^2$, although in practice this
aspect can be neglected. 

Figure \@ref(fig:preeclampsiaMA) also shows the results of a random 
effects analysis, together with the estimated Higgins' $I^2$ and 
heterogeneity variance $\tau^2$.


### Cumulative meta-analysis

A cumulative meta-analysis plot shows how evidence has accumulated 
over time. The $i$-th line in a cumulative meta-analysis plot 
is the summary produced by a meta-analysis of the 
first $i$-trials. 

```{r}
library(metadat)
```

:::{.example #bcg}
The dataset `dat.bcg`from the `R`package `metadat`
gives the results from 13 studies examining the effectiveness
of the Bacillus Calmette-Guerin (BCG) vaccine
against tuberculosis. 
:::
The meta-analysis can be performed with the function 
`meta::metabin()`: 

```{r, fig.height=7.5, fig.width=11, echo = TRUE}
meta1 <- metabin(event.e = tpos, n.e = tpos + tneg,
event.c = cpos, n.c = cpos + cneg, data=dat.bcg, sm = "RR",
method = "Inverse", studlab = paste(author, year))
### Forest plot
forest(meta1)
```
The cumulative meta-analysis can be performed with the function 
`meta::metacum()`: 

```{r, fig.height=7.5, fig.width=11, echo = TRUE}
forest(metacum(meta1, pooled="random"))
```


### Meta-regression
A meta-regression tries to explain heterogeneity with study-specific 
covariates. The visualisation can be done with bubble plots. 

:::{.example #bcg name="continued"}
In the `dat.bcg` dataset, the covariable `ablat` gives the absolute 
latitude of the study location (in degrees), and can possibly be
a relevant moderator of the effectiveness of the vaccine. 
:::

The meta-regression can be performed with the function 
`meta::metareg()`: 
```{r, echo = TRUE}
print(mr1 <- metareg(x=meta1, formula="ablat"))
```

The function `meta::bubble()` produces the bubble plot. 

```{r, echo = TRUE}
par(las=1)
bubble(mr1, lwd = 2, col.line = "blue", studlab=TRUE, xlim=c(-5, 60),
       cex.studlab=0.7, ylim=c(0.15, 1.8))
```



## Reporting bias

Reporting bias occurs when the publication of research results depends on their
nature and direction.
Sources of reporting bias include publication bias (where studies with negative or null findings may not be published due to decisions by researchers, referees, editors, or journals), language bias, citation bias, time lag bias, and outcome reporting bias (where outcomes are selectively reported, often due to changes in the research plan). These biases can lead to false conclusions and potentially harm patients.


### Funnel plot

A funnel plot is a scatter plot of a measure of study size, usually the (reversed) standard error, against the estimated treatment effects from individual studies. The funnel plot for the preeclampsia data is shown in Figure \@ref(fig:preeclampsiaFunnel). 

The smaller the study size, the wider the spread of the treatment effects and vice versa. If there is no bias, the point cloud has the form of a funnel (symmetrical). In contrast, it is asymmetrical 
if there is bias. However, it is an explorative tool and there is no quantitative information on the amount or the source of the bias.
Funnel plots and tools for meta-analysis in `R` are provided by the packages `meta` and `rmeta`.



```{r preeclampsiaFunnel, fig.cap = "Funnel plot for preeclampsia data from Example \\@ref(exm:preeclampsia)", echo=FALSE}
par(las = 1)
funnel(meta1)
``` 



:::{.example #tobacco}
@hackshaw1998 conducted a meta-analysis of 37 studies of the effect of environmental tobacco smoke on the risk of lung cancer in lifetime non-smokers. Spouses of smokers and non-smokers have been compared in terms of log relative risk. Are the results shown in a funnel plot in Figure \@ref(fig:smokingFunnel) affected by publication bias?
:::




```{r smokingFunnel, fig.cap = "Funnel plot of results in the meta-analysis about risk of lung cancer from passive smoking by @hackshaw1998.", echo=FALSE}
ps <- read.table("data/passive_smoking.txt", header=TRUE)
mymeta <- metagen(TE=lnRR, seTE=selnRR, data=ps)
par(las=1)
funnel(mymeta, xlab="log odds ratio")
``` 

Smaller studies tend to show greater effects, so there is some asymmetry. However, only visual inspection is not enough to decide if it is real bias or due to chance.


### The trim-and-fill method

The trim-and-fill method is based on the key assumptions that
publication bias is the reason of funnel plot asymmetry and that the studies with negative findings are suppressed.
The idea of this method is to:

1.  Trim-off the "asymmetric" side of a funnel plot, after estimating the number of studies in this group.
2.  Use the symmetric remainder to estimate the "true center".
3.  Impute trimmed studies and their missing "counterparts" around the center.
4.  Estimate $\theta$ and its variance based on the "filled" funnel plot.


:::{.example #stroke}
Figure \@ref(fig:trimfillStroke) illustrates the trim-and-fill method for 11 studies of the effect of using gangliosides in reducing case fatality and disability in acute ischaemic stroke.
:::


```{r trimfillStroke, echo=FALSE, fig.cap="Trim-and-fill method in Example \\@ref(exm:stroke). The black circles represent the 11 observed studies whereas the red studies on the left are imputed, based on symmetrically mirroring the trimmed studies on the right.", out.width='\\textwidth'}
knitr::include_graphics("figures/fig-trimfill05.pdf")
```

```{r ch8, echo=FALSE, results='hide'}
getL <- function(s,n){
  return( (4*s - n*(n+1))/(2*n-1))
}
library(xtable)
library(meta) # load the library
gs <- as.matrix(read.table("data/ganglioside.txt", header = TRUE))
n <- nrow(gs)
randomTE1 <- summary(metagen(gs[,1], gs[,2]))$random$TE
randomSE1 <- summary(metagen(gs[,1], gs[,2]))$random$seTE
print(randomTE1)
centered <- gs[,1]-randomTE1
gs <- cbind(gs, centered)
ranks <- rank(abs(gs[,3]))*sign(gs[,3])
gs <- cbind(gs, ranks)
s <- sum(ranks[ranks > 0])
L1 <- getL(s,n)
# omit 4 rightmost
randomTE2 <- summary(metagen(gs[-(8:11),1], gs[-(8:11),2]))$random$TE
print(randomTE2)
centered <- gs[,1]-randomTE2
gs <- cbind(gs, centered)
ranks <- rank(abs(gs[,5]))*sign(gs[,5])
gs <- cbind(gs, ranks)
s <- sum(ranks[ranks > 0])
L2 <- getL(s,n)
# omit 5 rightmost
randomTE3 <- summary(metagen(gs[-(7:11),1], gs[-(7:11),2]))$random$TE
centered <- gs[,1]-randomTE3
gs <- cbind(gs, centered)
ranks <- rank(abs(gs[,7]))*sign(gs[,7])
gs <- cbind(gs, ranks)
s <- sum(ranks[ranks > 0])
L3 <- getL(s,n)
# fill data
gs <- rbind(gs, NA, NA, NA, NA, NA)
filledlnOR <- 2*randomTE3 - gs[(7:11),1]
filledSE <- gs[(7:11),2]
gs <- cbind(gs,c(gs[1:11,1],filledlnOR), c(gs[1:11,2],filledSE))
sum_filled <- summary(metagen(gs[,9], gs[,10]))
randomTE_filled <- sum_filled$random$TE
randomSE_filled <- sum_filled$random$seTE
gs <- xtable(gs, digits=c(0,rep(2,3),0, 2,0,2,0,2,2))
```



Trim-and-fill with `R`:

```{r ch10, echo=TRUE}
library(meta) # load the library
gs <- as.matrix(read.table("data/ganglioside.txt", header = TRUE))
tf_gs <- trimfill(x = gs[,1], seTE = gs[,2], 
                  ma.fixed = FALSE, type = "L", silent = TRUE)
```     
        
```{r ch11, echo=TRUE}
print(tf_gs)
```


```{r trimfill06, results='hide', include=FALSE, fig.width=11, fig.height=6}
library(gplots)
par(mar=c(6,7,1,1), mgp=c(5,1.5,0), cex.axis=2, cex.lab=2.5, las=1)
d <- as.matrix(read.table("data/data2_hackshaw.txt", header = TRUE))
funnel(d[,1],d[,2], xlim=c(-1.5,1.5), pch=19, 
       xlab="log odds ratio", ylab="Standard error", 
       cex=2, level=0.95, yaxis="sei")
```

```{r trimfill07, results='hide', include=FALSE, fig.width=11, fig.height=8}
meta_d <- metagen(d[,1],d[,2], sm="RR")
sum_meta_d <- summary(meta_d)
trim_d <- trimfill(meta_d, ma.fixed=F, type="L")
sum_trim_d <- summary(trim_d)
layout(matrix(c(1,2),2,1), heights=c(3,1))
par(mar=c(6,7,1,1), mgp=c(5,1.5,0), cex.axis=2, cex.lab=2.5, las=1)
funnel(trim_d, pch=ifelse(trim_d$trimfill, 16, 16), cex=2,col=ifelse(trim_d$trimfill, 2, 1),
            level=0.95, comb.fixed=FALSE, xlab="log odds ratio", ylab="Standard error", yaxis="se")
plotCI(sum_meta_d$random$TE, 0, ui=sum_meta_d$random$upper, li=sum_meta_d$random$lower, cex=2, bty="n", xlim=c(-3,3), 
     xlab="Effect size", err="x", pch=23, lwd=2, yaxt="n", ylim=c(-1,1), 
    ylab="", sfrac=0.03, gap=0.8, pt.bg=1)
plotCI(sum_trim_d$random$TE, -0.5, ui=sum_trim_d$random$upper, li=sum_trim_d$random$lower, cex=2, bty="n", xlim=c(-3,3), 
     xlab="Effect size", err="x", pch=23, lwd=2, yaxt="n", ylim=c(-1,1), 
    ylab="", sfrac=0.03, gap=0.8, add=TRUE, pt.bg=2, col=2) 
```

:::{.example #tobacco name="continued"}
Data from the passive smoking example is illustrated in Figure \@ref(fig:smokingFunnel2).
Results from the trim-and-fill method are shown in Figure \@ref(fig:trimfillSmoking). The summary effect with (in red) and without (in black) the results from the imputed studies is shown at the bottom.
:::


```{r smokingFunnel2, fig.cap = "Funnel plot in the passive smoking Example \\@ref(exm:tobacco).", echo=FALSE, warning=FALSE}
library(meta)
ps <- read.table("data/passive_smoking.txt", header = TRUE) 
funnel(ps$lnRR,ps$selnRR,level=0.95, comb.fixed=TRUE, pch=19, 
       xlab="log relative risk", ylab="standard error")
```



```{r, echo=FALSE}
library(knitr)
library(dplyr)

# Assuming meta_d, sum_meta_d, and trim_d are already defined in the R environment
data <- data.frame(
  Model = c("Observed", "Filled"),
  `Number of studies` = c(meta_d$k, trim_d$k),
  OR = c(round(exp(sum_meta_d$random$TE), 2), round(exp(sum_trim_d$random$TE), 2)),
  `95% CI` = c(
    paste0("(", round(exp(sum_meta_d$random$lower), 2), ", ", round(exp(sum_meta_d$random$upper), 2), ")"),
    paste0("(", round(exp(sum_trim_d$random$lower), 2), ", ", round(exp(sum_trim_d$random$upper), 2), ")")
  )
)

kable(
  data,
  col.names = c("Model", "Number of studies", "OR", "95% CI"),
  escape = FALSE,
  booktabs = TRUE, align = "l"
)
```


```{r trimfillSmoking, echo=FALSE, fig.cap="Trim-and-fill method in the passive smoking Example \\@ref(exm:tobacco).", fig.align='center', out.width='\\textwidth'}
knitr::include_graphics("figures/fig-trimfill07.pdf")
```



### Tests for funnel plot asymmetry

If visual inspection of a funnel plot is not enough, statistical tests can be 
performed. 
Popular methods are Begg's rank correlation method or Egger's test (weighted regression).

#### Begg's rank correlation test {-}
Begg's rank correlation test computes standardized treatment estimates

\begin{equation*}
\theta_i^*=\frac{\theta_i-\hat{\theta}}{\sqrt{v_i^*}},
\end{equation*}

where $\hat{\theta}$ is the fixed-effect estimate of the summary effect and 
$v_i^*=v_i - 1/(\sum v_j^{-1})$ is the variance of $\theta_i-\hat{\theta}$.
Then, it tests the null hypothesis that *Kendell's rank correlation* 
(Kendell's $\tau$) between $\theta_i^*$ and $v_i^*$ is zero. This test 
may have low power if the number of studies is small.

A rank correlation test can be performed using `R` as shown below for Example \@ref(exm:tobacco).

```{r ch14a, echo=TRUE, results='markup'}
head(ps)
metabias(ps$lnRR, ps$selnRR, method="Begg")
``` 

  

#### Egger's weighted regression {-}

The idea is to perform weighted regression of $\hat \theta_i$ on $s_i$,

\begin{equation*}
\hat \theta_i= a + b \cdot s_i + \mbox{error},
\end{equation*}

with weights $w_i = 1/s_i^2$. The slope $b$ is a measure of bias.

:::{.example #tobacco name="continued"}
Egger's weighted regression can be performed in `R` as shown below for the passive smoking example:
:::

```{r ch15, echo=TRUE, results='markup'}
theta <- ps$lnRR
s <- ps$selnRR
egger1 <- lm(theta ~ s, weights = 1/s^2)
```

```{r results="asis"}
library(biostatUZH)
knitr::kable(tableRegression(egger1, latex = FALSE, xtable = FALSE))
``` 

Figure \@ref(fig:smokingEgger) shows a visualization of Egger's weighted regression in the passive smoking example.
  

```{r smokingEgger, fig.cap = "Visualization of Egger's weighted regression in the passive smoking example.", echo=FALSE}
theta.hat <- mean(theta, weight=w)
par(las=1)
plot(s, theta, ylim=c(-0.3,1.25), pch=16, xlab="standard error", ylab="log relative risk")
abline(egger1, col=2)
lines(s,predict(egger1, interval="confidence")[,2], col=2, lty=2)
lines(s,predict(egger1, interval="confidence")[,3], col=2, lty=2)
lines(c(0,1), rep(theta.hat, 2), col=3, lty=1)
legend("topleft", col=c(2,2,3), lty=c(1,2,1), legend=c("Fit", "95%-CI", "Fixed effect estimate"), cex=0.9)
``` 


Egger's test in `R`:

```{r ch14, echo=TRUE, results='markup'}
metabias(ps$lnRR, ps$selnRR, method="Egger")
``` 

 

#### Limit meta-analysis {-}

The aim of a limit meta-analysis is to compute a bias-adjusted overall 
effect estimate by extrapolating the weighted regression to 
$s_i \to 0$. 

A limit meta-analysis can be performed in `R`
using the function `metasens::limitmeta()`. 
For Example \@ref(exm:tobacco), this gives:
```{r echo = TRUE}
psMeta <- metagen(TE=ps$lnRR, seTE=ps$selnRR)
library(metasens)
limitmeta(psMeta)
```



## Additional references

See @bland (Chapter 17) and @matthews (Chapter 12).




