--- 
title: "Clinical Biostatistics"
author: "Leonhard Held, with contributions from Charlotte Micheloud, Lisa Hofer, Stefanie von Felten, Samuel Pawel"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: bib.bib  
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "Based on the lecture notes from STA404: Clinical Biostatistics."
always_allow_html: yes
---

\newcommand{\eg}{\abk{\latin{e.\,g}}} 
\usepackage[textsize=scriptsize]{todonotes}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xifthen}
\usepackage{ifthen}
\usepackage{dsfont} % for the indicator function
\newcommand{\given}{\,\vert\,} 
\def\SE{\mbox{se}}
\def\EF{\mbox{EF}}
\newcommand{\LOA}{\mbox{LoA}}
\newcommand{\LOAlow}{\mbox{LoA}_{\tiny \mbox{low}}}
\newcommand{\LOAup}{\mbox{LoA}_{\tiny \mbox{up}}}
\def\pobs{p_\mathrm{obs}}
\def\pexp{p_\mathrm{exp}}
\DeclareMathOperator{\Nor}{N} % Normal
\newcommand{\abs}[1]{\left\lvert#1\right\rvert} % Absolutbetrag
\newcommand{\Ind}{\mathsf{I}} % Indikatorfunktion

% Operatoren
\renewcommand{\P}{\operatorname{\mathsf{Pr}}} % probability (P was already def.)
\DeclareMathOperator{\E}{\mathsf{E}} % expectation
\DeclareMathOperator{\Var}{Var} % variance
\DeclareMathOperator{\se}{se} % standard error
\DeclareMathOperator{\Cov}{Cov} % covariance
\DeclareMathOperator{\Cor}{Cor} % correlation
\DeclareMathOperator{\sign}{sign} % signum
\DeclareMathOperator{\logit}{logit} % logit fct.
\DeclareMathOperator{\expit}{expit} % expit fct.
\DeclareMathOperator{\Mod}{Mod} % mode
\DeclareMathOperator{\Med}{Med} % median
\DeclareMathOperator{\diag}{diag} % diagonal matrix
\DeclareMathOperator{\trace}{tr} % trace
\DeclareMathOperator{\Bin}{Bin} % Binomial

\newcommand{\partials}[3][]{
  \ifthenelse{\isempty{#1}}
  {\frac{d\,#2}{d\,#3}}
  {\frac{d^{#1} #2}{d\,#3^{#1}}}
} 

\def\PE{\mbox{PE}}
\def\OR{\mbox{OR}}
\def\HR{\mbox{HR}}
\def\RR{\mbox{RR}}
\def\EF{\mbox{EF}}
\def\RV{\mbox{RV}}
\def\RD{\mbox{RD}}
\def\ARR{\mbox{ARR}}
\def\RRR{\mbox{RRR}}
\def\NNT{\mbox{NNT}}
{\widehat{#2}_{\scriptscriptstyle{ML}}}
\newcommand{\Deff}{D$_{\small{eff}}$ % Varianz
\newcommand{\sima}{\mathrel{\overset{a}{\thicksim}}} % approx. Vtlg.

# Preface {-}

<div style="text-align: right;">
"Medicine is a science of uncertainty and an art of probability." <br>
William Osler (1849-1919).
</div>

Biostatistics is a fundamental discipline at the core of modern health
data science [@lee_biostatistics_2019]. It is the science of
managing medical uncertainty and biostatistical methods play a key
role in the scientific assessment of the main areas of medical
practice:

- Diagnosis
- Aetiology
- Treatment
- Prognosis


<div style="text-align: right;">
  Leonhard Held
  
  `r Sys.Date()`
</div>











```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      cache = FALSE, 
                      fig.align = 'center')

knitr::opts_chunk$set(
  todo = function(text) {
    if (knitr::is_html_output()) {
      sprintf('<div class="todo">%s</div>', text)
    } else if (knitr::is_latex_output()) {
      sprintf('\\todo{%s}', text)
    } else {
      text
    }
  }
)
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
